Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis

### 2020-11-26 manuscript/competing_risks.Rmd
Version as sent for approval to co-authors

---
title: "Survival - frequentist"
output: html_notebook
---

```{r}
devtools::load_all()
library(tidyverse)
library(survival)

@@

# Competing risks death vs. discharged

```{r}
data_surv <- wide %>%  
  filter(day > 0) %>%
  mutate(daym1 = day - 1,
    status = fct_collapse(breathing, Hospitalized = breathing_levels) %>% factor(ordered = FALSE) %>% fct_relevel("Hospitalized", "Discharged", "Death") %>% fct_explicit_na("Hospitalized")
  ) %>%
  inner_join(data$patient_data, by = c("hospital_id", "patient_id"))

f_competing_hcq <- Surv(daym1, day,status) ~ age + sex + took_hcq

#survcheck(f_competing_hcq, data = data_surv, id = patient_id)
fit_competing_hcq <- coxph(f_competing_hcq, data = data_surv, id = patient_id)
fit_competing_hcq

@@

hypothesis_res_list[["competing_hcq"]]
```

```{r}
f_competing_hcq_az <- update.formula(f_competing_hcq, . ~ . + took_az)
fit_competing_hcq_az <- coxph(f_competing_hcq_az, data = data_surv, id = patient_id)

@@

hypothesis_res_list[["competing_hcq_az"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az", multiplier = -1),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az,
                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, hcq"),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az,
                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq", multiplier = -1)
  )
hypothesis_res_list[["competing_hcq_az"]]
```

```{r}
f_competing_hcq_az_fpv <- update.formula(f_competing_hcq_az, . ~ . + took_favipiravir)
fit_competing_hcq_az_fpv <- coxph(f_competing_hcq_az_fpv, data = data_surv, id = patient_id)

@@

hypothesis_res_list[["competing_hcq_az_fpv"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_death, fit_competing_hcq_az_fpv,
                                        coefficient_name = "took_favipiravirTRUE", transition = "1:3", adjusted = "age, sex, hcq, az", model_check = "Problematic"),
  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_hospital, fit_competing_hcq_az_fpv,
                                        coefficient_name = "took_favipiravirTRUE", transition = "1:2", adjusted = "age, sex, hcq, az", model_check = "Problematic", multiplier = -1)
  )
hypothesis_res_list[["competing_hcq_az_fpv"]]
```

```{r}
f_competing_hcq_az_admitted <- update.formula(f_competing_hcq_az, . ~ . + admitted_for_covid)
fit_competing_hcq_az_admitted <- coxph(f_competing_hcq_az_admitted, data = data_surv, id = patient_id)

@@

hypothesis_res_list[["competing_hcq_az_admitted"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_admitted,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, admitted, az"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_admitted,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, admitted, az", multiplier = -1),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az_admitted,
                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, admitted, hcq"),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az_admitted,
                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, admitted, hcq", multiplier = -1)
  )
hypothesis_res_list[["competing_hcq_az_admitted"]]
```

```{r}
f_competing_hcq_az_hospital <- update.formula(f_competing_hcq_az, . ~ . + hospital_id)
data_surv_hospital <- data_surv %>% 
  mutate(hospital_id = if_else(hospital_id %in% c("DJkhZ", "lRBHa", "nDINR"), "small", hospital_id))
fit_competing_hcq_az_hospital <- coxph(f_competing_hcq_az_hospital, data = data_surv_hospital, id = patient_id)
fit_competing_hcq_az_hospital

hypothesis_res_list[["competing_hcq_az_hospital"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, hospital"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, hospital", multiplier = -1),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, hcq, hospital"),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq, hospital", multiplier = -1)
  )
hypothesis_res_list[["competing_hcq_az_hospital"]]

@@

```{r}
f_competing_hcq_az_comorbidities_sum <- update.formula(f_competing_hcq_az, . ~ . + comorbidities_sum)
fit_competing_hcq_az_comorbidities_sum <- coxph(f_competing_hcq_az_comorbidities_sum, data = data_surv, id = patient_id)
fit_competing_hcq_az_comorbidities_sum

hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_comorbidities_sum,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, comorbidities (sum)"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_comorbidities_sum,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, comorbidities (sum)", multiplier = -1)
  )
hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]]
```

# Peak IL-6, peak D-dimer

```{r}
f_competing_marker_base <- Surv(daym1, day,status) ~ age + sex
f_competing_il_6 <- update.formula(f_competing_marker_base, . ~ . + log_peak_IL_6)

@@

hypothesis_res_list[["competing_il_6"]] <-
  frequentist_hypothesis_res_from_coxph(hypotheses$IL_6_death, fit_competing_il_6,
                                        coefficient_name = "log_peak_IL_6", transition = "1:3", adjusted = "age, sex", model_check = "Suspicious")
hypothesis_res_list[["competing_il_6"]]
```
```{r}
f_competing_d_dimer <- update.formula(f_competing_marker_base, . ~ . + log_peak_d_dimer)

@@

hypothesis_res_list[["competing_d_dimer"]] <-
  frequentist_hypothesis_res_from_coxph(hypotheses$d_dimer_death, fit_competing_d_dimer,
                                        coefficient_name = "log_peak_d_dimer", transition = "1:3", adjusted = "age, sex", model_check = "Problematic")
hypothesis_res_list[["competing_d_dimer"]]
```

>>>>

---
title: "Survival - frequentist"
output: bookdown::pdf_document2
---

```{r setup_competing, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = FALSE)
devtools::load_all()
library(tidyverse)
library(survival)

@@

# Competing risks proportional hazards models

In all of those models we use the `coxph` function from the `Survival` package and treat death and discharged as competing risks. The `status` column in the data corresponds to the three-valued state "Hospitalized" (1) / "Discharged" (2) / "Death" (3). I.e. in all model outputs below, there are two transitions "1:2" is the transition to "Discharged" and "1:3" is the transition to "Death". We evaluate the hypotheses about treatment effects or marker associations directly via the estimated log hazard ratio of the treatment for the relevant transition. Notably, the sign of the effect on the transition to "Discharged" is flipped for the multiverse analysis because it assumes that positive effect means increased hospital stay.

We do not work with models with more than about 5 predictors, as those become impossible to estimate due to insufficient amount of events.

```{r}
data_surv <- wide %>%  
  filter(day > 0) %>%
  mutate(daym1 = day - 1,
    status = fct_collapse(breathing, Hospitalized = breathing_levels) %>% factor(ordered = FALSE) %>% fct_relevel("Hospitalized", "Discharged", "Death") %>% fct_explicit_na("Hospitalized")
  ) %>%
  inner_join(data$patient_data, by = c("hospital_id", "patient_id")) 
```

## Treatment effects

### Age + sex + hydroxychloroquine

The model uses the formula:

```{r}
f_competing_hcq <- Surv(daym1, day, status) ~ age + sex + took_hcq
f_competing_hcq
```
Fitted coefficients below:

```{r}

fit_competing_hcq <- coxph(f_competing_hcq, data = data_surv, id = patient_id)
fit_competing_hcq

@@

#hypothesis_res_list[["competing_hcq"]]
```

### Age + sex + hydroxychloroquine + azithromycin

The formula becomes:

```{r}
f_competing_hcq_az <- update.formula(f_competing_hcq, . ~ . + took_az)
f_competing_hcq_az
```

The fitted coefficients are:

```{r}

fit_competing_hcq_az <- coxph(f_competing_hcq_az, data = data_surv, id = patient_id)

@@

hypothesis_res_list[["competing_hcq_az"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az", multiplier = -1),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az,
                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, hcq"),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az,
                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq", multiplier = -1)
  )
#hypothesis_res_list[["competing_hcq_az"]]
```

### Age + sex + hydroxychloroquine + azithromycin + favipiravir

The formula is:

```{r}
f_competing_hcq_az_fpv <- update.formula(f_competing_hcq_az, . ~ . + took_favipiravir)
f_competing_hcq_az_fpv
```

And the fitted model is:

```{r}

fit_competing_hcq_az_fpv <- coxph(f_competing_hcq_az_fpv, data = data_surv, id = patient_id)

@@

hypothesis_res_list[["competing_hcq_az_fpv"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_death, fit_competing_hcq_az_fpv,
                                        coefficient_name = "took_favipiravirTRUE", transition = "1:3", adjusted = "age, sex, hcq, az", model_check = "Problematic"),
  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_hospital, fit_competing_hcq_az_fpv,
                                        coefficient_name = "took_favipiravirTRUE", transition = "1:2", adjusted = "age, sex, hcq, az", model_check = "Problematic", multiplier = -1)
  )
#hypothesis_res_list[["competing_hcq_az_fpv"]]
```

### Age + sex + hydroxychloroquine + azithromycin + admitted

The formula is:

```{r}
f_competing_hcq_az_admitted <- update.formula(f_competing_hcq_az, . ~ . + admitted_for_covid)
f_competing_hcq_az_admitted
```

The fitted coefficients are:

```{r}
fit_competing_hcq_az_admitted <- coxph(f_competing_hcq_az_admitted, data = data_surv, id = patient_id)

@@

hypothesis_res_list[["competing_hcq_az_admitted"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_admitted,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, admitted, az"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_admitted,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, admitted, az", multiplier = -1),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az_admitted,
                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, admitted, hcq"),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az_admitted,
                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, admitted, hcq", multiplier = -1)
  )
#hypothesis_res_list[["competing_hcq_az_admitted"]]
```

### Age + sex + hydroxychloroquine + azithromycin + hospital

The new formula is:

```{r}
f_competing_hcq_az_hospital <- update.formula(f_competing_hcq_az, . ~ . + strata(hospital_id))
f_competing_hcq_az_hospital
```

The resulting coefficients are:

```{r}

data_surv_hospital <- data_surv %>% 
  mutate(hospital_id = if_else(hospital_id %in% c("DJkhZ", "lRBHa", "nDINR"), "small", hospital_id))
fit_competing_hcq_az_hospital <- coxph(f_competing_hcq_az_hospital, data = data_surv_hospital, id = patient_id)
fit_competing_hcq_az_hospital

hypothesis_res_list[["competing_hcq_az_hospital"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, hospital"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, hospital", multiplier = -1),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, hcq, hospital"),
  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq, hospital", multiplier = -1)
  )
#hypothesis_res_list[["competing_hcq_az_hospital"]]

@@

### Age + sex + hydroxychloroquine + azithromycin + comorbidities

The formula becomes:

```{r}
f_competing_hcq_az_comorbidities_sum <- update.formula(f_competing_hcq_az, . ~ . + comorbidities_sum)
f_competing_hcq_az_comorbidities_sum
```

Model fit below:

```{r}
fit_competing_hcq_az_comorbidities_sum <- coxph(f_competing_hcq_az_comorbidities_sum, data = data_surv, id = patient_id)
fit_competing_hcq_az_comorbidities_sum

hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_comorbidities_sum,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, comorbidities (sum)"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_comorbidities_sum,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, comorbidities (sum)", multiplier = -1)
  )
#hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]]
```

## Marker effects

### Age + sex + IL-6

The formula is:

```{r}
f_competing_marker_base <- Surv(daym1, day,status) ~ age + sex
f_competing_il_6 <- update.formula(f_competing_marker_base, . ~ . + log_peak_IL_6)
f_competing_il_6
```

where `log_peak_IL_6` is the logarithm of the peak IL-6 value seen up to the given time. The fitted model is:

```{r}

@@

hypothesis_res_list[["competing_il_6"]] <-
  frequentist_hypothesis_res_from_coxph(hypotheses$IL_6_death, fit_competing_il_6,
                                        coefficient_name = "log_peak_IL_6", transition = "1:3", adjusted = "age, sex", model_check = "Suspicious")
#hypothesis_res_list[["competing_il_6"]]
```
Note the warning on convergence and the huge `se` for the `log_peak_IL_6` coefficient for the 1:2 transition. We mark this model as "Suspicious".

### Age + sex + D-dimer

The formula is:

```{r}
f_competing_d_dimer <- update.formula(f_competing_marker_base, . ~ . + log_peak_d_dimer)
f_competing_d_dimer
```

where `log_peak_d_dimer` is the logarithm of the peak D-dimer value seen up to the given time. The fitted model is:

```{r}

@@

hypothesis_res_list[["competing_d_dimer"]] <-
  frequentist_hypothesis_res_from_coxph(hypotheses$d_dimer_death, fit_competing_d_dimer,
                                        coefficient_name = "log_peak_d_dimer", transition = "1:3", adjusted = "age, sex", model_check = "Suspicious")
#hypothesis_res_list[["competing_d_dimer"]]
```
We see similar, but even more extreme convergence issues as before, we mark the model "Suspicious".

### 2020-11-26 manuscript/hmm.Rmd
Version as sent for approval to co-authors

```{r setup_hmm, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = FALSE)

@@

We treat the breathing support required as a Markov process, either with states directly corresponding to the breathing support and directly observable (Fig. \@ref(fig:hmmstates)a), or with the states carrying an unobserved binary improving/worsening component which is unobservable (Fig. \@ref(fig:hmmstates)b) while the breathing support dimension still being completely observable. The former completely observable model is called _simple_ in the following text while the latter is called _complex_. This is unlike many uses of HMMs where the observation process is described by a general observation matrix. The reason is that the breathing support actually used is very likely to correspond to the breathing support needed by the patient. In initial versions of our models we also tested models that treated the breathing support observed as a noisy realisation of the actual state, but the fitted probabilities of such imprecise observations were always very low, so we ended up not using them.

@@

Models will be presented in a unified structure that we explain with the first model and won't be repeated

@@

For this model, the $\log(R_{i,j})$ elements are then modeled via the `brms` formula

@@

Our code is setup so that the formula can combine predictors drawn from the description of rates (see Table \@ref(tab:statessimple)) and those taken from patient's data (both the fixed baseline characteristics and time-varying values). There is a separate coefficient for each rate and then the other predictors act uniformly on each rate group. The `took_XXX` variables are boolean predictors that are set to true for all days since the first dose of the treatment was given - i.e. we assume the treatment alters the overall disease progression irreversibly.

@@

And the resulting log ORs:

@@

```{r}
trt_group_m4_formula <-  update.formula(trt_group_m3_formula,  ~ . + (comorbidities_sum || rate_group)) 
trt_group_m4_formula
```

@@

```{r}
trt_rate_m4_formula <-  update.formula(trt_rate_m3_formula, ~ . + (comorbidities_sum || .rate_id))
trt_rate_m4_formula
```

@@

```{r}
d_dimer_m2_formula <-  update.formula(d_dimer_m1_formula, ~ . + (0 + age_norm + is_male || rate_group))
d_dimer_m2_formula
```

@@

```{r}
IL_6_m2_formula <-  update.formula(IL_6_m1_formula, ~ . + (0 + age_norm + is_male || rate_group))
IL_6_m2_formula
```

@@

```{r}
trt_group_m4_formula
```

>>>>

```{r setup_hmm, echo=FALSE, message=FALSE, warning = FALSE, results = "hide",cache=FALSE}
knitr::opts_chunk$set(echo = FALSE)

@@

We treat the breathing support required as a Markov process, either with states directly corresponding to the breathing support and directly observable (Fig. \@ref(fig:hmmstates)a), or with the states carrying a binary improving/worsening component which is unobservable (Fig. \@ref(fig:hmmstates)b) while the breathing support dimension still being completely observable. The former completely observable model is called _simple_ in the following text while the latter is called _complex_. This results in a very restricted observation matrix. The reason is that the breathing support actually used is very likely to correspond to the breathing support needed by the patient. In initial versions of our models we also tested models that treated the breathing support observed as a noisy realisation of the actual state, but the fitted probabilities of such imprecise observations were always very low, so we ended up not using them.

@@

Models will be presented in a unified structure that we explain with the first model and will not be repeated

@@

For this model, the $\log(R_{i,j})$ elements are modeled via the `brms` formula

@@

Our code is setup so that the formula can combine predictors drawn from the description of rates (see Table \@ref(tab:ratessimple)) and those taken from patient's data (both the fixed baseline characteristics and time-varying values). There is a separate coefficient for each rate and then the other predictors act uniformly on each rate group. The `took_XXX` variables are boolean predictors that are set to true for all days since the first dose of the treatment was given - i.e. we assume the treatment alters the overall disease progression irreversibly.

@@

And the resulting log ORs/log duration ratios:

@@

```{r}
trt_group_m4_formula <-  update.formula(trt_group_m3_formula,  ~ . + (comorbidities_sum || rate_group)) 
trt_group_m4_formula
```
Where `comorbidities_sum` is a comorbidity score as described in Section \@ref(comorbiditiessum).

@@

```{r}
trt_rate_m4_formula <-  update.formula(trt_rate_m3_formula, ~ . + (comorbidities_sum || .rate_id))
trt_rate_m4_formula
```
Where `comorbidities_sum` is a comorbidity score as described in Section \@ref(comorbiditiessum).

@@

```{r}
d_dimer_m2_formula <-  update.formula(d_dimer_m1_formula, ~ . + (0 + age_norm + is_male || rate_group))
d_dimer_m2_formula
```
Where `comorbidities_sum` is a comorbidity score as described in Section \@ref(comorbiditiessum).

@@

```{r}
IL_6_m2_formula <-  update.formula(IL_6_m1_formula, ~ . + (0 + age_norm + is_male || rate_group))
IL_6_m2_formula
```
Where `comorbidities_sum` is a comorbidity score as described in Section \@ref(comorbiditiessum).

@@

```{r}
trt_group_m4_formula
```
Where `comorbidities_sum` is a comorbidity score as described in Section \@ref(comorbiditiessum).

### 2020-11-26 manuscript/jm.Rmd
Version as sent for approval to co-authors

---
title: "Joint modelling"
output: html_document
---

@@

```{r setup}
devtools::load_all()
library(tidyverse)
library(survival)
library(rstanarm)
library(tidybayes)

@@

@@

#data_d_dimer %>% filter(d_dimer == 0)
# data_d_dimer %>% group_by(sex, ever_hcq) %>% summarise(pts = length(unique(patient_id)), n_dimer = sum(!is.na(d_dimer)))
fit_d_dimer_age_sex <- stan_jm_with_cache(formulaLong = log_d_dimer ~ day + (1  | patient_id),
              dataLong = data_d_dimer %>% filter(!is.na(d_dimer)),
              formulaEvent = Surv(daym1, day, event_death) ~ age + sex,
              dataEvent = data_d_dimer,
              id_var = "patient_id",
              time_var = "day",
              cache_file = paste0(cache_dir, "/jm_d_dimer_age_sex.rds"))

@@

#length(unique(data_d_dimer$patient_id))
# to_pred <- data_d_dimer %>% group_by(patient_id) %>% summarise(n_d_dimer = sum(!is.na(d_dimer))) %>%
#   filter(n_d_dimer > 1) %>% pull(patient_id)
#data_d_dimer %>% filter(patient_id == unique(data_d_dimer$patient_id)[31])

@@

jm_hypothesis_res_list[["d_dimer_age_sex"]]
```

@@

fit_d_dimer_age_sex_hcq

#length(unique(data_d_dimer$patient_id))
# to_pred <- data_d_dimer %>% group_by(patient_id) %>% summarise(n_d_dimer = sum(!is.na(d_dimer))) %>%
#   filter(n_d_dimer > 1) %>% pull(patient_id)
#data_d_dimer %>% filter(patient_id == unique(data_d_dimer$patient_id)[31])

@@

jm_hypothesis_res_list[["d_dimer_age_sex_hcq"]]
```

```{r}
# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[1:11])
# plot(p1, plot_observed = TRUE)                     
# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[12:14])
# plot(p1, plot_observed = TRUE)                     
# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[15:18])
# plot(p1, plot_observed = TRUE)                     
# 
# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[19:20])
# plot(p1, plot_observed = TRUE)                     
# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[21:22])
# plot(p1, plot_observed = TRUE)                     
# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[22:26])
# plot(p1, plot_observed = TRUE)
# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[27:29])
# plot(p1, plot_observed = TRUE)                     
# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[30:32])
# plot(p1, plot_observed = TRUE)                     
```

@@

jm_hypothesis_res_list[["IL_6_age_sex"]]

@@

@@

jm_hypothesis_res_list[["IL_6_age_sex_hcq"]]

@@

```{r}
for(id in unique(data_IL_6$patient_id)) {
  p1 <- posterior_traj(fit_IL_6_age_sex_hcq, ids = id)
  print(plot(p1, plot_observed = TRUE)                     )
}

>>>>

---
title: "Joint modelling"
output: bookdown::pdf_document2
bibliography: Covid19.json
---

@@

Joint models have two components:

- _longitudinal model_ describes the evolution of some quantity/quantities over time (in our case a biological marker)
- _time-to-event model_ describes the risk of event over time assuming proportional hazards. This risk can be affected by the value of the marker.

Modelling those two jointly means that the marker values are imputed between measurements according to the longitudinal model, but information also flows in the other direction. We use the implemention from the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5] - consult the reference for an introduction to this class of models.

We use those models only for association between markers (IL-6 and D-dimer) and disease progression. Those analyses ended up not being reported in the main paper as the results are inconclusive.

We evaluate the associations directly via the coefficient describing the association between the longitudinal and time-to-event component. We focus solely on the "Death" event.

```{r setup_jm, cache=FALSE, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(cache = TRUE, echo=FALSE)

devtools::load_all()
library(tidyverse)
library(survival)
library(rstanarm)
library(tidybayes)

@@

## D-dimer

### Age + sex

The longitudinal formula is `log_d_dimer ~ day + (1  | patient_id)` and the time-to-event formula is `Surv(daym1, day, event_death) ~ age + sex`. The fitted coefficients are:

@@

fit_d_dimer_age_sex <- stan_jm_with_cache(formulaLong = log_d_dimer ~ day + (1  | patient_id),
              dataLong = data_d_dimer %>% filter(!is.na(d_dimer)),
              formulaEvent = Surv(daym1, day, event_death) ~ age + sex,
              dataEvent = data_d_dimer,
              id_var = "patient_id",
              time_var = "day",
              cache_file = paste0(cache_dir, "/jm_d_dimer_age_sex.rds"))

@@

@@

#jm_hypothesis_res_list[["d_dimer_age_sex"]]
```

### Age + sex + hydroxychloroquine

The time-to-event formula is expanded to `Surv(daym1, day, event_death) ~ age + sex + took_hcq`. The fitted coefficients are:

@@

fit_d_dimer_age_sex_hcq
```

@@

#jm_hypothesis_res_list[["d_dimer_age_sex_hcq"]]
```

The problems in the model are visible when visualising the fitted longitudinal trajectories for some patients: there are very few measurements and thus huge uncertainties:

```{r, warning = FALSE, message = FALSE}
 p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[23:28])
 plot(p1, plot_observed = TRUE)                     
```

## IL-6

### Age + sex

The longitudinal formula is `log_IL_6 ~ day + (1 + day | patient_id)` and the time-to-event formula is `Surv(daym1, day, event_death) ~ age + sex`. The fitted coefficients are:

@@

#jm_hypothesis_res_list[["IL_6_age_sex"]]

@@

### Age + sex + hydroxychloroquine

The time-to-event formula becomes `Surv(daym1, day, event_death) ~ age + sex + took_hcq`. The fitted model is:

@@

#jm_hypothesis_res_list[["IL_6_age_sex_hcq"]]

@@

Similarly, we can explore the posterior trajectories, showing similar issues as before: too wide uncertainty.

```{r, message = FALSE}
 p1 <- posterior_traj(fit_IL_6_age_sex_hcq, ids = unique(data_IL_6$patient_id)[c(8,14,23,13,6,1)])
 plot(p1, plot_observed = TRUE)                     
```

### 2020-11-26 manuscript/manuscript.Rmd
Version as sent for approval to co-authors

---
title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    number_sections: false
    reference_docx: "output_template.docx"
abstract: |  
  **Abstract:** We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. Patients in the dataset were treated with hydroxychloroquine (N = 108), azithromycin (N = 72), favipiravir (N = 9), convalescent plasma (N = 7), dexamethasone (N = 4) and remdesivir (N = 3), often in combination. Most patients were admitted during the first wave of the epidemic. To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.  Weak evidence to constrain the potential efficacy of azithromycin and favipiravir can be extracted from the data. Additionally we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset.
bibliography: Covid19.json
csl: journal-of-medical-virology.csl
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---

>>>>

---
title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    number_sections: false
    reference_docx: "output_template.docx"
abstract: |  
  **Abstract:** We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. Patients in the dataset were treated with hydroxychloroquine (N = 108), azithromycin (N = 72), favipiravir (N = 9), convalescent plasma (N = 7), dexamethasone (N = 4) and remdesivir (N = 3), often in combination. Most patients were admitted during the first wave of the epidemic. To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.  Weak evidence to constrain the potential efficacy of azithromycin and favipiravir can be extracted from the data. Additionally, we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset.
bibliography: Covid19.json
csl: journal-of-medical-virology.csl
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---
